---
title: "Time series analysis: basics"
author: "Steven Moran & Marco Maiolini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
      toc: true
bibliography: '../2_writing_scientific_reports/references.bib'
---

# Overview

See the [lecture slides](Time_Series.pdf) and take the DataCamp tutorial for deeper insights and discussion.

Let's dive into two case studies using time series analysis. Both datasets are available in the [data](data/) directory.


# Case study: COVID-19

[COVID-19](https://en.wikipedia.org/wiki/COVID-19) data is publicly available.

Load the R packages.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(tseries)
```

We have some data about COVID-19. Let's load it.

```{r, message=FALSE, warning=FALSE}
data <- read_csv("data/Covid_19_cases_EU.csv")
```

And take a look.

```{r}
head(data)
str(data)
```

Maybe we want to explore it a bit.

```{r}
table(data$Countries)

data %>% filter(Countries == "Sweden")
```
Let's formulate a hypothesis.

* H0: The covid cases will decrease (not increase)
* H1: The covid cases will increase

***

Although the loaded data is not a time series object in R, the raw data is time series. Convert the data into time series.

```{r}
is.ts(data)
class(data)
data$Date <- as.Date(data$Date)
data$Cases <- as.numeric(data$Cases)
data.ts <- ts(data$Cases, start = c(2020, 2), end = c(2022, 2), frequency = 365)
is.ts(data.ts)
print(data.ts)
str(data.ts)
start(data.ts)
end(data.ts)
frequency(data.ts)
deltat(data.ts)
```

Visualize it as a time-series.

```{r}
# par (mfrow=c(2,1), mar=c(4,4,4,4))
ts.plot(data.ts, col = 1:4, xlab = "Date", ylab = "Cases", main = "Covid cases Sweden", type = "l")
```

ACF plot.

```{r}
acf(data.ts, main = "acf plot of Covid cases in Sweden")
```

Which model is more suitable?

```{r}
data_diff <- diff(data.ts)

par(mfrow=c(1,1), mar=c(4,4,4,4))

ts.plot(data_diff, xlab = "Date", ylab = "Cases", main = "WN Covid cases --", type = "l")

arima(x= data.ts, order = c(0,0,0)) # Fit WN model

arima(x= data.ts , order = c(0,1,0))  # Fit RW model

data_AR <- arima(data.ts, order = c(1,0,0))
print(data_AR)
ts.plot(data.ts, xlab = "Date", ylab = "Cases", main = "AR Covid cases ---", type = "l")
AR_fitted <- data.ts - residuals(data_AR)
points(AR_fitted, type = "l", col = 2, lty = 2)

data_MA <- arima(data.ts, order = c(0,0,1))
print(data_AR)
ts.plot(data.ts, xlab = "Date", ylab = "Cases", main = "MA Covid cases ---", type = "l")
MA_fitted <- data.ts - residuals((data_MA))
points(MA_fitted, type = "l", col = 2, lty = 2)
```

Let's forecast the time series trends.

```{r}
ts.plot(data.ts, xlab = "Date", ylab = "Cases", main = "Prediction Covid cases ---", type = "l")
AR_forecast <- predict(data_AR, n.ahead = 10)$pred
AR_forecast_se <- predict(data_AR, n.ahead = 10)$se
points(AR_forecast, type = "l", col = 2)
points(AR_forecast - 2*AR_forecast_se, type = "l", col = 2, lty = 2)
points(AR_forecast + 2*AR_forecast_se,type = "l", col = 2, lty = 2)

ts.plot(data.ts, xlab = "Date", ylab = "Cases", main = "Prediction Covid cases ---", type = "l")
MA_forecast <- predict(data_MA, n.ahead = 10)$pred
MA_forecast_se <- predict(data_MA, n.ahead = 10)$se
points(MA_forecast, type = "l", col = 2)
points(MA_forecast - 2*MA_forecast_se, type = "l", col = 2, lty = 2)
points(MA_forecast + 2*MA_forecast_se,type = "l", col = 2, lty = 2)
```

Which model is better?

```{r}
cor(AR_fitted, MA_fitted)
```

Autoregression.

```{r}
AIC(data_AR) # Better
BIC(data_AR)
```

Moving Average.

```{r}
AIC(data_MA) # Better
BIC(data_MA)
```



# Case study: Species in Italy

* H0: Different species have different trend
* H1: Both species have similar trend

Load the R libraries.

```{r}
library(tidyverse)
library(tseries)
```

Load the data

```{r}
data <- read.csv("data/Species_Individuals_center_Italy.csv")
```

Have a look.

```{r}
str(data)
head(data)
table(data$Species)
```

Turn the data into a time series object.

```{r}
data <- data %>%
 mutate (Date = paste(Year, Month, '1', sep = '-'))

data$Date <- as.Date(data$Date, format = '%Y-%m-%d')

data$Individuals <- as.numeric(data$Individuals)

str(data)

data0 <- filter(data, Species == "Sus scrofa")

is.ts(data0)

data0

data0.ts <- ts(data0$Individuals, start = c(2020,10),frequency = 12)

is.ts(data0.ts)

print(data0.ts)

str(data0.ts)

start(data0.ts)

end(data0.ts)

frequency(data0.ts)

deltat(data0.ts)
```


Plot the time series (plot + acf).

```{r}
#Axis control
xmin<-min(data0$Date,na.rm=T);xmax<-max(data0$Date,na.rm=T) #ESTABLISH X-VALUES (MIN & MAX)
ymin<-min(data0$Individuals,na.rm=T);ymax<-max(data0$Individuals,na.rm=T) #ESTABLISH Y-VALUES (MIN & MAX)
xseq<-seq.Date(xmin,xmax,by='1 month') #CREATE DATE SEQUENCE THAT INCREASES BY MONTH FROM DATE MINIMUM TO MAXIMUM
yseq<-round(seq(0,ymax,by=10),0) # CREATE SEQUENCE FROM 0-350 BY 50

#Plot
plot(Individuals~Date,data=data0,type='b',ylim=c(0,ymax),axes=F,xlab='',ylab='', main = "Individuals presence");box() #PLOT LEVEL AS A FUNCTION OF DATE, REMOVE AXES FOR FUTURE CUSTOMIZATION 
axis.Date(side=1,at=xseq,format='%Y-%m',labels=T,las=3) #ADD X-AXIS LABELS WITH "YEAR-MONTH" FORMAT
axis(side=2,at=yseq,las=2) #ADD Y-AXIS LABELS
mtext('Date',side=1,line=5) #X-AXIS LABEL
mtext('Individuals',side=2,line=4) #Y-AXIS LABEL
```

```{r}
#Remove NAs
mean(data0.ts, na.rm = TRUE)
data0.ts[8] <- mean(data0.ts, na.rm = TRUE)

#Plot without NAs
# par (mfrow=c(2,1), mar=c(4,4,4,4))

plot(data0.ts, type = "b",xlab = "Date", ylab = "N individuals", main = "Boar individals", axes = T);box()
```


```{r}
#axis(2);
#axis.Date(1,xlim=xseq, at=xseq, format='%Y-%m', labels=T)
acf(data0.ts, main = "acf boar spotted")
```

Remove NAs.

```{r}
mean(data0.ts, na.rm = TRUE)
data0.ts[8] <- mean(data0.ts, na.rm = TRUE)
```

Plot without NAs

```{r}
# par (mfrow=c(2,1), mar=c(4,4,4,4))
plot(data0.ts, type = "b",xlab = "Date", ylab = "N individuals", main = "Boar individals", axes = T);box()
```

```{r}
#axis(2);
#axis.Date(1,xlim=xseq, at=xseq, format='%Y-%m', labels=T)
acf(data0.ts, main = "acf boar spotted")
```

Which model is more suitable?

```{r}
data0_AR <- arima(data0.ts, order = c(1,0,0))
print(data0_AR)
ts.plot(data0.ts, xlab = "Date", ylab = "Individuals", main = "AR Individuals boar", type = "b")
AR_fitted0 <- data0.ts - residuals(data0_AR)
points(AR_fitted0, type = "l", col = 2, lty = 2)

data0_MA <- arima(data0.ts, order = c(0,0,1))
print(data0_MA)
ts.plot(data0.ts, xlab = "Date", ylab = "Individuals", main = "MA Individuals boar", type = "b")
MA_fitted0 <- data0.ts - residuals((data0_MA))
points(MA_fitted0, type = "l", col = 2, lty = 2)
```

Which model is better?

```{r}
cor(AR_fitted0, MA_fitted0)
```

Autoregression

```{r}
AIC(data0_AR) #better
BIC(data0_AR)
```

Moving Average

```{r}
AIC(data0_MA) #better
BIC(data0_MA)
```

Now two compare a second species.

Filter [Capreolus capreolus](https://en.wikipedia.org/wiki/Roe_deer).

```{r}
data1 <- filter(data, Species == "Capreolus capreolus")

is.ts(data1)

data1 <- data1 %>%
  mutate (Date = paste(Year, Month, '1', sep = '-'))

data1$Date <- as.Date(data1$Date, format = '%Y-%m-%d')

data1

data1$Individuals <- as.numeric(data1$Individuals)

data1.ts <- ts(data1$Individuals, start = c(2020, 10), end = c(2021, 09), frequency = 12)

is.ts(data1.ts)

print(data1.ts)

str(data1.ts)

start(data1.ts)

end(data1.ts)

frequency(data1.ts)

deltat(data1.ts)
```

Plot as a time-series (plot + acf).

```{r}
#Axis control
xmin<-min(data1$Date,na.rm=T);xmax<-max(data1$Date,na.rm=T) #ESTABLISH X-VALUES (MIN & MAX)
ymin<-min(data1$Individuals,na.rm=T);ymax<-max(data1$Individuals,na.rm=T) #ESTABLISH Y-VALUES (MIN & MAX)
xseq<-seq.Date(xmin,xmax,by='1 month') #CREATE DATE SEQUENCE THAT INCREASES BY MONTH FROM DATE MINIMUM TO MAXIMUM
yseq<-round(seq(0,ymax,by=10),0) # CREATE SEQUENCE FROM 0-350 BY 50
#Plot
plot(Individuals~Date,data=data1,type='b',ylim=c(0,ymax),axes=F,xlab='',ylab='', main = "Individuals presence");box() #PLOT LEVEL AS A FUNCTION OF DATE, REMOVE AXES FOR FUTURE CUSTOMIZATION 
axis.Date(side=1,at=xseq,format='%Y-%m',labels=T,las=3) #ADD X-AXIS LABELS WITH "YEAR-MONTH" FORMAT
axis(side=2,at=yseq,las=2) #ADD Y-AXIS LABELS
mtext('Date',side=1,line=5) #X-AXIS LABEL
mtext('Individuals',side=2,line=4) #Y-AXIS LABEL
```

Remove NAs.

```{r}
#mean(data1.ts, na.rm = TRUE)
#data1.ts[8] <- mean(data.ts, na.rm = TRUE)
```

Plot.

```{r}
# par (mfrow=c(2,1), mar=c(4,4,4,4))
plot(data1.ts, type = "b",xlab = "Date", ylab = "N individuals", main = "Roe deer individals", axes = T);box()
```

```{r}
#axis(2);
#axis.Date(1,xlim=xseq, at=xseq, format='%Y-%m', labels=T)
acf(data1.ts, main = "acf roe deer spotted")
```

Which model is more suitable?

```{r}
data1_AR <- arima(data1.ts, order = c(1,0,0))
print(data1_AR)
ts.plot(data1.ts, xlab = "Date", ylab = "Individuals", main = "AR Individuals roe deer", type = "b")
AR_fitted1 <- data1.ts - residuals(data1_AR)
points(AR_fitted1, type = "l", col = 2, lty = 2)
```

```{r}
data1_MA <- arima(data1.ts, order = c(0,0,1))
print(data1_MA)
ts.plot(data1.ts, xlab = "Date", ylab = "Individuals", main = "MA Individuals roe deer", type = "b")
MA_fitted1 <- data1.ts - residuals((data1_MA))
points(MA_fitted1, type = "l", col = 2, lty = 2)
```

Which model is better?

```{r}
cor(AR_fitted1, MA_fitted1)
# ts.plot(AR_fitted1, MA_fitted1)
```

Autoregression.

```{r}
AIC(data1_AR) # Better
BIC(data1_AR) # Better
```

Moving Average.

```{r}
AIC(data1_MA)
BIC(data1_MA)
```


Final plot.

```{r}
ts.plot(data0.ts, xlab = "Date", ylab = "Individuals", main = "AR Individuals boar", type = "b")
AR_fitted0 <- data0.ts - residuals(data0_AR)
points(AR_fitted0, type = "l", col = 2, lty = 2)
```


```{r}
ts.plot(data1.ts, xlab = "Date", ylab = "Individuals", main = "AR Individuals roe deer", type = "b")
AR_fitted1 <- data1.ts - residuals(data1_AR)
points(AR_fitted1, type = "l", col = 2, lty = 2)
```






# Helpful hints

Install and load the appropriate R libraries.

```{setup, eval = FALSE}
# install.packages("tidyverse")
# install.packages("tseries")
library(tidyverse)
library(tseries)
```

Load and check your data.

Explore the dataset and filter if necessary to have only one object (as showed in the COVID or species example).

Create your time series object.

```{Ts object, eval = FALSE}
data$Date <- as.Date(data$Date) #Format as date

data$Cases <- as.numeric(data$Cases) # Format as number

data.ts <- ts(data$Cases, start = c(2020, 2), end = c(2022, 2), frequency = 365)

is.ts(data.ts)
```

* Plot your time series and the Autocorrelation plot
* Fit your time series object in the models that we saw in class
* Using the Autoregressive and the Simple moving average models try to forecast how your time series will develop in the future.
* Which is the best model for your data?
* Evaluate your results, there are limits using time series models? Motivate your answer.


# Where to find more data

Here are some other sources of time series data:

- Epidemiology - https://www.ecdc.europa.eu/en/covid-19/data
- Finance - https://unibit.ai/solution (It requires a login)
- Demography - https://fred.stlouisfed.org/
- Miscellaneous - https://db.nomics.world/
- Species presence - https://www.inaturalist.org/
- R dataset - type "Nile", "eu_stocks" or "AirPassengers" in R



